Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-09-05 doc/papers/cgo2014/paper.tex
Include BC's review

    As more scientists use these scripting languages, there is an on-going
    research effort to compile them into efficient native code. However as
    these languages favor a high-level coding style, the computation kernels are
    different from what we use to see in traditional C or Fortran programs:
    explicit raw loops are replaced by high-level constructs and calls to the
    relevant toolbox/module/library/package. As a consequence, traditional
    algorithms to perform
    loop optimization techniques like fusion, tiling and related
    vectorization or parallelization techniques are not directly applicable.

@@

The Python language~\cite{rossum97} has a rich ecosystem that now provides a
full toolkit to carry out scientific experiments, from core scientific routines
with the Numpy package\cite{oliphant2007,numpyarray2011}, to scientific
packages with the Scipy package~\cite{scipy}, plotting facilities with the
Matplotlib package, enhanced terminal and notebooks with IPython. As a
consequence, there has been a move from historical languages like Fortran to
Python, as showcased by the success of the Scipy conference.

As Python based scientific tools gets widely used, the question of High
performance Computing naturally arises, and it is the focus of many recent
research. Indeed, although there is a great gain in productivity when using
these tools, there is also a performance gap that needs to be filled.

This paper focuses on compilation techniques that are relevant for the
optimization of high-level numerical kernels written in Python using the Numpy
package. Its introduces the implementation of these techniques in the Pythran
compiler and provides a benchmarks test suite for performance measures along
several Python compilers.
Section~\ref{sec:stackoverflow} presents a case study based on user
input gathered on the StackOverflow questions and answers website.
Section~\ref{sec:optimize} focuses on a simple yet representative kernel and
showcases the optimization opportunities typically found in Numpy-based
kernels. Existing compilation approaches to optimize such kernels are discussed
in Section~\ref{sec:compilers}. Section~\ref{sec:pythran} emphasises on a
compiler-runtime cooperative approach to solve the performance issues.
Finally, experimentations showcasing important performance gains are detailed in
Section~\ref{sec:xp}.

@@

One of the first steps required when designing an optimizing compiler is to
gather enough test cases, or benchmarks, in order to drive the optimization
process toward realistic examples. To achieve this goal, we have gathered a few
examples from existing compilers' test suite, but these may have been modified to
overcome some compiler limitations. A less biased source of benchmarks should
be independent from the compiler community and focused on matching user needs.
To achieve that goal, we collected synthetic scientific kernels from the
StackOverflow website.

StackOverflow~\footnote{\url{http://stackoverflow.com/}} is a question and
answer site for programmers. Any registered user can submit questions and other
users may answer. A voting system is then used to sort the answers. Said
otherwise, the website provides a database of questions and answers on
programming topics, including Python.

@@

For instance, question \href{http://stackoverflow.com/questions/7741878}{7741878}
proposes to use an explicit loop to iteratively compute the $L^2$ norm of each
row of a matrix, as illustrated in Listing~\ref{lst:l2norm-loopy}, and one of
the proposed answer is to use the more compact version presented in
Listing~\ref{lst:l2norm}. This answer no longer uses any explicit loops and
roughly achieves a $\times 4$ speedup over the loop version.

@@

Newcomers to Python with a prior C or Fortran experience are likely to make
use of raw loops similar to Listing~\ref{lst:l2norm-loopy}, but as their
knowledge of the Numpy API and of good programming practices grows, they get to 
write code similar to Listing~\ref{lst:l2norm}. 
The fact that in Numpy, high-level constructs run faster is also a strong 
motivation. As a direct consequence, a compiler should focus on optimizing
high-level constructs rather than raw
loops. Both to encourage good practice with respect to Numpy code, and because
this is the kind of code Numpy users ultimately write.

The benchmarks collected during this process are publicly available on
\url{https://github.com/serge-sans-paille/numpy-benchmarks}, with a set of
benchmarking routine specialized for major Python/Numpy compilers.

@@

Let us consider the kernel illustrated in Listing~\ref{lst:rosen} and adapted
from the Scipy source code of \texttt{scipy.optimize.rosen}. This kernel makes
use of Numpy's \texttt{sum} function, Python's square notation and Numpy's
array slicing. It is a good example of high-level Python kernel, although
using the function \texttt{scipy.optimize.rosen} directly would naturally make sense.

This section goes through all the optimization opportunities in
Listing~\ref{lst:rosen} as a showcase of what an optimizing compiler could do.

In Numpy, all point-to-point array operations allocate a new array that holds
the computation result. This behavior is consistent with many Python standard
module, but it is a very inefficient design choice, as it keeps on polluting
the cache with potentially large fresh storage and add an extra
allocation/deallocation costly operation. In the \textit{rosen} function from
Listing~\ref{lst:rosen}, 7 temporary arrays are allocated (slicing does not
create a temporary array but a view) to hold intermediate steps. Had the
computation been lazy, no temporary would have been needed.

As Numpy is a native library mostly written in C, each operator computation is
performed by a function implemented as a loop performing a single operation,
and the operator chaining is done at the interpreter level. This is a typical
problem in library design: if only a small set of functions is provided, it
prevents the optimization of merging multiple operators into a single specialized
operator, and provided many operator combinations as part of the library yields
better performance to the price of API bloat. The Listing~\ref{lst:rosen}
illustrates the use of a small set of functions: a loop is used for
each temporary computation, plus an extra loop for the \texttt{numpy.sum}
reduction, where a single loop would have been needed with operator fusion.

Without operator fusion, there would be very little benefit to generate SIMD
instructions for the respective array operations used by each operator, as the
memory load and store would have dominated the execution time. This is even more important as
Numpy typically operates on double precision floats, which means only two (SSE)
to four (AVX) scalar per vector registers.

Parallelization also suffers from the lack of operator fusion: a
synchronization fence is needed between each temporary computation.  Hence the
loop computation intensity would be very low compared to the memory pressure
implied by two array reads and one array write for a single binary operator.

On the opposite, if all computations were merged into a single loop using
temporaries elimination and operator fusion, parallelization would be more
effective as barriers between binary operations are no longer needed and loads
and stores will be used to perform multiple operations.

%%
%%
\section{Existing Compilation Approach}
\label{sec:compilers}

@@

\subsection{Cython}
%
Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
syntax with typing information, calls to native functions from third party C
libraries, and a limited set of parallelism constructs, such as the possibility
to define parallel loops, but no task parallelism. When possible, it unboxes % ref?
Python variables to improve performance. Without type annotations, the
performance improvement is not significant, but given enough type annotation,
Cython can generates code that runs as fast as its C equivalent.

The drawback is that the programmer has to avoid high-level constructs and
instead expresses his algorithm using raw loops.  The \textit{pairwise}
function rewritten for Cython is shown listing~\ref{lst:pairwise-cython},
showing usage of compiler annotations in the form of function decorators.

@@

The Numba compiler uses additional type information collected at run-time to
generate sequential LLVM bytecode.  Numba uses Just-In-Time(JIT) compilation
and does not suffer from backward-incompatibility issues: a single
\texttt{@jit} decorator is used to flag kernels to optimize. When meeting an
unsupported construct, Numba silently falls back to Python C-API calls.

Like Python, Numba is providing its best speedup on code with raw loops, which
is not encouraging programmers to express their code using the higher-level
construct available with Numpy. However, when input constraint are met, it
exhibits performance close to Cython, or even better when conversion to GPU
code is possible.

@@

Parakeet~\cite{parakeet2012} follows an approach similar to Numba, that is JIT
compilation of selected kernels. It uses C annotated with OpenMP directives as
a target language, but limits its scope to the Numpy package, only supporting
the subset of the Python language required to use Numpy functions.
Additionally, it supports implicit parallelism using an implicit mapping
between Numpy functions and a set of parallel primitives including
\texttt{map}, \texttt{scan}, and \texttt{reduce}.  As it applies optimizations
such as operator fusion, it encourages programmer to express their code using
the higher-level construct available with Numpy.

@@

Copperhead~\cite{copperhead2011} is a functional, data parallel language
embedded in Python. It uses $n$-uplets, Numpy arrays and lists as its core data
structure and prohibits usage of many control-flow operators such as loops,
enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
intrinsics to exhibit parallelism. But it can be efficiently compiled to either
CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.
\texttt{http://thrust.github.com/}} library. Like Numba and Parakeet, it uses
Python decorators to identify hot-spots that are JIT-compiled to native code.
As it requires to write functional code, it is not compatible with the Numpy
interface, although several Numpy functions could be rewritten in terms of
Copperhead intrinsics.

@@

@@

There is no need to reinvent the wheel when compiling numerical Python kernels.
A profitable approach is to map Python high-level to existing ones, or
translate them into existing them.\todo{Mehdi: the previous sentence makes no sense to me}
As a consequence, this section focuses on the translation of Python constructs
to efficient C++ constructs. The choice of
C++ instead of C is motivated by the opportunities offered by meta programming
and the interesting similarities between C++11~\cite{isocxx11} features and
Python constructs.
The meta-programming part of C++ makes it more than a traditional compiler
runtime, closer to a code generator backend.

@@

Many C++ linear algebra libraries~\cite{eigen,ublas,esterie2014} use \emph{expression
templates}~\cite{expression_templates, et2012} as a way to efficiently describe high
level matrix expressions while avoid the creation of intermediate storage.
Additionally, it has the same effect as loop fusion, solving the issues
described in \S~\ref{sec:temporaries-elimination} and
\S~\ref{sec:operator-fusion}.

@@

Note that turning such Numpy expressions into expression templates is
equivalent to converting from a strict evaluation to a non-strict, or lazy,
evaluation. This is always legal in the context of Numpy because none of the
computations considered for expression template conversion have side effect.

@@

Let us illustrate the algorithm on a simple case. Python code source and the
initial effect graphs are illustrated on Figure~\ref{fig:argument-effects}. The
only node with write effects is \texttt{bar} that modifies \texttt{n} through a
subscript. Outgoing edges (1) and (2) can be removed because they are
outgoing edges of \texttt{bar} starting by \texttt{n}. Incoming edge (0) ends
by \texttt{n} that is written by \texttt{bar}, so it propagates the write
effects to the source node corresponding to the function \texttt{root}. After
removing this edge, the only remaining edges are propagating read and the
analysis ands. The analysis informs that both \texttt{root} update \texttt{r}
and \texttt{bar} update \texttt{n}, but \texttt{foo} doesn't update its
argument.

@@

            \draw[->] (bar) -- node[right] {$(\texttt{n,s})$} node[left] {(2)} (foo);
            \draw[->] (bar.east) node[right, yshift=-1em] {(1)} -| ($(bar.east) + (.5,.75)$) node[right, yshift=-.8em] {$(\texttt{n,n})$} -| (bar.north) ;
            \draw[->] (root) -- node[right, yshift=.8em] {$(\texttt{r,n})$} node[left, yshift=.8em] {(0)} (bar);
        \end{tikzpicture}
        \caption{Initial argument effect graph..}
        \label{fig:argument-effects-graph}
    \end{subfigure}

@@

A naive translation of such call to a generic C++ function would generate a
call to a C++ function that would allocate enough memory to hold three integers,
fill the corresponding buffer and return it. A smarter version would generate a
static array holding the relevant values. Figure~\ref{fig:unroll-range}
compares two translations of a Python snippet that iterates over a fixed-size
list. The Clang compiler successfully turns Listing~\ref{lst:stat} into a
single constant. It also statically computes the return value for
Listing~\ref{lst:dyna}, but still generates the initialization code for the
\texttt{std::vector}. The GCC compiler fails to statically compute the return
value in both cases. This example shows that constant propagation should be
done at a higher level than C++, and so be part of the Python optimizer.

@@

Once interprocedural constant evaluation is done (eventually followed by
forward substitution), iteration over fixed size lists may be exposed, giving an
opportunity to fully unroll these loops at the Python level, thus avoiding the
suboptimal C++ translation observed for Listing~\ref{lst:stat}
and~\ref{lst:dyna}.

@@

By construction of the expression template, the \texttt{eval} method used in
\texttt{from.eval(i)} does not have side effects, and the Python to C++
compiler can assert there is no shared memory between \texttt{to} and any
component of \texttt{from}. In that case, all loops can be made parallel, and
the innermost loop is eventually vectorizable.

@@

A conservative approach is to only start a parallel region if the loop trip
count is large enough. We avoid by using the \texttt{if} clause for parallel
regions to do so in order to prevent the unnecessary overhead. Instead a manual
if is used: the loop statement is duplicated and guarded by the appropriate
check, where only one loop holds the parallel annotations.
\todo{this paragraph seems to be about a particular runtime issue, this shouldn't matter.
Moreover, how is it relevant to this paper? It is an "implementation detail"}

Note that this parallelization scheme is compatible with the manual
parallelization of Python programs proposed in~\cite{pyhpc2013}.
\todo{I think a specific section could be made on this topic}

@@

The ideas described in this paper are implemented in the Pythran
compiler~\cite{pythran2013}, a translator from a subset of Python to
C++11~\cite{isocxx11}. The architecture of the compiler is briefly described
in Figure~\ref{fig:pythran-compiler}. The input is a Python module written in
the Pythran subset. Pythran translates it in its internal representation, a
simplified Python AST. It performs various optimization then outputs either
Python code, in a source-to-source fashion, or C++ template code that makes
call to the pythonic library that typically implements the \texttt{ndarray}
interface. User annotations can be used to instantiate this code for the proper
types and generate a native library. This library relies on Boost.Python
library to match Python's C API.

The Pythran compiler is an open source project publicly released under the BSD
license~\footnote{\url{http://pythonhosted.org/pythran/}}.

@@

As introduced in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test 
suite to evaluate the optimizations described in\S~\ref{sec:pythran}. 
Pythran is used (commit: 488b7ba) with and without SIMD optimization. 
The results are compared with Pypy (commit: 632c832bd32f), Parakeet (commit: 4e55810), 
Numba (commit: b8a79c5) and Python 2.7.

numpy-benchmarks run 11 times benchmarks with an execution time around 1 seconde.
THe best time, average time and standard deviation is return from these 11 runs.

Benchmarking is done on two architectures with different backend compilers.

The first one holds 8 AMD Magny Cours processors for a
total of 48 cores. Each core has access to 64\,KB of L1 cache,
512\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 6 cores of a processor. This configuration
provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
up to SSE4. We refer to this configuration as \textbf{AMD48}.
On AMD48, the backend compilers is the clang-omp
version from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to
benefits from OpenMP improvements. This Clang version is based on Clang 3.4
release (LLVM:233b1e3f034, clang-omp:6f358bcd87).

The second one holds one Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
8 cores (multithreading is not used). Each node can access to 64\,KB of L1
cache, 256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 8 cores. This configuration provides a total of 64\,GB of 
main memory. It supports up to AVX. We refer to this configuration as
\textbf{Intel8}.
On Intel8, backend compilers is GCC 4.9 (20140528) with libgomp.

@@

% For performance comparison between different Pythran version
To compare the different optimization schemes, we run the numpy-benchmarks test
suite multiple times on the \textit{rosen} benchmark. The algorithm is the one
on Listing~\ref{lst:rosen} and input is a raw array of 1,000,000 double elements.
Figures~\ref{fig:rosenamd48} and \ref{fig:rosenintel8} illustrate the results.
The speedup difference between the two architectures can be explain by AVX support on
the Intel8 architecture while AMD48 supports up to SSE4 only.
We can also notice importance of forward substitution on this benchmarks as
we avoid extra loop, 2 temporary array assignment and avoid some load/store for
SIMD instructions.

@@

@@

Table~\ref{tbl:overview} presents the average execution time for each kernel
and each compiler, with the lowest values highlighted in
bold font. Zeros are used when the given compiler failed to
compile the kernel. Concerning the Numba compiler, this
benchmarks conﬁrms that it works best on explicit loops. It
gives non performance gain when run on high-level kernels,
either falling back to the generation of calls to Python C
API, or stopping compilation. PyPy is the most robust of
the compilers, but to a few exceptions, it does not generate
code more efﬁcient than the original version. This reﬂects
the fact that Numpy support is still a work-in-progress task.
Parakeet performs extremely well on high-level benchmarks.
Although it does not supports some Numpy constructs, it
successfully parallelize many of them, and also generates
efﬁcient code for explicit loops. Pythran still performs better
than Parakeet on some kernels, whenever it successfully
vectorizes operations.

\begin{table*}

@@

>>>>

    As more scientists use these scripting languages, there is an on-going
    research effort to compile them into efficient native code. However these
    languages favor a high-level coding style, so the computation kernels are
    different from those used in traditional C or Fortran programs:
    explicit raw loops are replaced by high-level constructs and calls to the
    relevant toolbox/module/library/package. As a consequence, traditional
    algorithms to perform loop optimization techniques like fusion, tiling and
    related vectorization or parallelization techniques are not directly
    applicable.

@@

The Python language~\cite{rossum97} has a rich ecosystem that now provides a
full toolkit to carry out scientific experiments, from core scientific routines
with the Numpy package\cite{oliphant2007,numpyarray2011}, to scientific
packages with Scipy~\cite{scipy}, plotting facilities with the
Matplotlib package, enhanced terminal and notebooks with IPython. As a
consequence, there has been a move from historical languages like Fortran to
Python, as showcased by the success of the Scipy conference.

As Python based scientific tools get widely used, the question of High
performance Computing naturally arises, and it is the focus of many recent
research. Indeed, although there is a great gain in productivity when using
these tools, there is also a performance gap that needs to be filled.

This paper focuses on compilation techniques that are relevant for the
optimization of high-level numerical kernels written in Python using the Numpy
package. Its introduces the implementation of these techniques in the Pythran
compiler and provides a benchmark test suite for performance measures along
several Python compilers.
Section~\ref{sec:stackoverflow} presents a case study based on user
input gathered from the StackOverflow questions and answers website.
Section~\ref{sec:optimize} focuses on a simple yet representative kernel and
showcases the optimization opportunities typically found in Numpy-based
kernels. Existing compilation approaches to optimize such kernels are discussed
in Section~\ref{sec:compilers}. Section~\ref{sec:pythran} emphasises on
compiler-runtime cooperation to solve the performance issues.
Finally, experimentations showcasing important performance gains are detailed in
Section~\ref{sec:xp}.

@@

One of the first steps required when designing an optimizing compiler is to
gather enough test cases, or benchmarks, in order to drive the optimization
process toward realistic examples. To achieve this goal, we have gathered a few
examples from existing compilers' test suites, but these may have been modified to
overcome some compiler limitations. A less biased source of benchmarks should
be independent from the compiler community and showed focus on matching user needs.
To achieve that goal, we collected synthetic scientific kernels from the
StackOverflow website.

StackOverflow\footnote{\url{http://stackoverflow.com/}} is a question and
answer site for programmers. Any registered user can submit questions and other
users may answer. A voting system is then used to sort the answers. Said
otherwise, the website provides a database of questions and answers on
programming topics, including Python.

@@

For instance, question \href{http://stackoverflow.com/questions/7741878}{7741878}
proposes to use an explicit loop to iteratively compute the $L^2$ norm of each
row of a matrix, as illustrated in Listing~\ref{lst:l2norm-loopy}, and one of
the proposed answer is to use the more compact version presented in
Listing~\ref{lst:l2norm}. This answer no longer uses any explicit loop and
roughly achieves a $\times 4$ speedup over the loop version.

@@

Newcomers to Python with a prior C or Fortran experience are likely to make
use of raw loops similar to Listing~\ref{lst:l2norm-loopy}, but as their
knowledge of the Numpy API and of good programming practices grows, they get to 
write code similar to Listing~\ref{lst:l2norm}. 
The fact that Numpy high-level constructs runs faster is also a strong 
motivation. As a direct consequence, a compiler should focus on optimizing
high-level constructs rather than raw
loops. Both to encourage good practice with respect to Numpy code, and because
this is the kind of code Numpy users ultimately write.

A selection of benchmarks collected during this process, selected for their
conciseness and representativity of high-level calls, are publicly available on
\url{https://github.com/serge-sans-paille/numpy-benchmarks}, with a set of
benchmarking routines specialized for major Python/Numpy compilers. This
benchmark is used in \S~\ref{sec:xp} to evaluate the validity of the approach
described in this paper.

@@

This section briefly presents the core concepts of the Numpy package, then goes
through all the optimization opportunities in a small kernel as a showcase of
the optimization opportunities.

@@

\subsection{The Rosenbrock Function}

This section uses the kernel illustrated in Listing~\ref{lst:rosen} and adapted
from the Scipy source code of \texttt{scipy.optimize.rosen}. It uses Numpy's
\texttt{sum} function, Python's square notation and Numpy's array slicing. It
is a good example of high-level Python kernel, although using the function
\texttt{scipy.optimize.rosen} directly would naturally make sense.

Note that due to dynamic typing, this function can take arrays of different
shapes and types as input.

In Numpy, any point-to-point array operation allocates a new array that holds
the computation result. This behavior is consistent with many Python standard
module, but it is a very inefficient design choice, as it keeps on polluting
the cache with potentially large fresh storage and adds extra
allocation/deallocation operations, that have a very bad caching effect. In the
\textit{rosen} function from Listing~\ref{lst:rosen}, 7 temporary arrays are
allocated (slicing does not create a temporary array but a view) to hold
intermediate steps. Had the expression been evaluated lazily, no temporary
would have been needed.

As Numpy is a native library mostly written in C, each operator computation is
performed by a function implemented as a loop performing a single operation,
and the operator chaining is done at the interpreter level. This is a typical
problem in library design: if only a small set of functions is provided, it
prevents the optimization of merging multiple operators into a single specialized
operator. On the contrary, providing many operator combinations as part of the library yields
better performance to the price of API bloat. Listing~\ref{lst:rosen}
illustrates the use of a small set of functions: a loop is generated for
each temporary computation, plus an extra loop for the \texttt{numpy.sum}
reduction, whereas a single loop would have been necessary with operator fusion.

Without operator fusion, there would be very little benefit to generate SIMD
instructions for the respective array operations used by each operator, as the
memory loads and stores would have dominated the execution time. This is even more important as
Numpy typically operates on double precision floats, which means only two (SSE)
to four (AVX) scalars per vector registers.

Parallelization would also suffer from the lack of operator fusion: a
synchronization fence is needed between each temporary computation.  Hence the
loop computation intensity would be very low compared to the memory pressure
implied by two array reads and one array write for a single binary operator.

On the opposite, if all computations were merged into a single loop using
temporaries elimination and operator fusion, parallelization would be more
effective as barriers between binary operations are no longer needed and loads
and stores are counterbalanced by several operations.

%%
%%
\section{Existing Compilation Approaches}
\label{sec:compilers}

\subsection{Cython}
%
Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
syntax with typing information, calls to native functions from third party C
libraries, and a limited set of parallelism constructs, such as the possibility
to define parallel loops, but no task parallelism. When possible, it unboxes
Python variables to improve performance, i.e. it removes the generic object
layer around a variable to turn it into a native type. Without type
annotations, the performance improvement is not significant, but given enough
type annotations, Cython can generate code that runs as fast as its C
equivalent.

The drawback is that the programmer has to avoid high-level constructs and
must instead express his algorithm using raw loops.  The \textit{pairwise}
function rewritten for Cython is shown in Listing~\ref{lst:pairwise-cython},
showing usage of compiler annotations in the form of function decorators.

@@

The Numba compiler collects additional type information at run-time to
generate sequential LLVM bytecode.  Numba uses Just-In-Time(JIT) compilation
and does not suffer from backward-incompatibility issues: a single
\texttt{@jit} decorator is provided to flag kernels to optimize. When meeting an
unsupported construct, Numba silently falls back to Python C-API calls.

Like Cython, Numba achieves its best speedups on codes with raw loops, which
does not encourage programmers to express their codes using the higher-level
constructs available within Numpy. However, when input constraints are met, it
exhibits performance close to Cython, or even better when conversion to GPU
code is possible.

@@

Parakeet~\cite{parakeet2012} follows an approach similar to Numba, that is JIT
compilation of selected kernels. It uses C annotated with OpenMP directives as
a target language, but limits its scope to the Numpy package, only supporting
the subset of the Python language required to use Numpy functions.
Additionally, it supports implicit parallelism using an implicit mapping
between Numpy functions and a set of parallel primitives including
\texttt{map}, \texttt{scan}, and \texttt{reduce}.  As it applies optimizations
such as operator fusion, it encourages programmer to express their code using
the higher-level constructs available within Numpy.

@@

Copperhead~\cite{copperhead2011} is a functional, data parallel language
embedded in Python. It accepts $n$-uplets, Numpy arrays and lists as its core data
structures and prohibits usage of many control-flow operators such as loops,
enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
intrinsics to exhibit parallelism. But it can be efficiently compiled to either
CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.
\texttt{http://thrust.github.com/}} library. Like Numba and Parakeet, it uses
Python decorators to identify hot-spots that are JIT-compiled to native code.
As it requires to write functional code, it is not compatible with the Numpy
interface, although several Numpy functions could be rewritten in terms of
Copperhead intrinsics.

@@

The ideas described in this paper are implemented in the Pythran
compiler~\cite{pythran2013}, a translator from a subset of Python to
C++11~\cite{isocxx11}. The architecture of the compiler is briefly described
in Figure~\ref{fig:pythran-compiler}. The input is a Python module written in
the subset accepted by the compiler. Pythran translates it in its internal representation, a
simplified Python AST. It performs various optimizations then outputs either
Python code, in a source-to-source fashion, or C++ templated code calling the pythonic library that typically implements the \texttt{ndarray}
interface. User annotations can be used to instantiate this code for the proper
types and generate a native library. This library relies on Boost.Python
library to match Python's C API.

The Pythran compiler is an open source project publicly released under the BSD
license\footnote{\url{http://pythonhosted.org/pythran/}}.

Pythran keeps the static compilation approach used by Cython---JIT compilation
is mainly used to statically type kernels before code generation by Numba and
parakeet. Unlike Cython, it maintains full Python backward compatibility.
Following the Parakeet approach, it focuses on high-level constructs, while
still generating efficient code for explicit loop and subscripts.

@@

There is no need to reinvent the wheel when compiling numerical Python kernels.
A profitable approach is to map Python high-level functions to existing ones.
As a consequence, this section focuses on the translation of Python constructs
to efficient C++ constructs. The choice of C++ instead of C is motivated by the
opportunities offered by meta programming and the interesting similarities
between C++11~\cite{isocxx11} features and Python constructs.  The
meta-programming part of C++ makes it more than a traditional compiler runtime,
closer to a code generator back-end.

@@

Many C++ linear algebra libraries~\cite{eigen,ublas,esterie2014} use \emph{expression
templates}~\cite{expression_templates, et2012} as a way to efficiently describe high
level matrix expressions while avoiding the creation of intermediate storage.
Additionally, it has the same effect as loop fusion, solving the issues
described in \S~\ref{sec:temporaries-elimination} and
\S~\ref{sec:operator-fusion}.

@@

Note that turning such Numpy expressions into expression templates is
equivalent to converting from a strict evaluation to a non-strict, or lazy,
evaluation. This is always legal in the context of Numpy because none of the
computations considered for expression template conversion have side effects.

@@

Let us illustrate the algorithm on a simple case. Python code source and the
initial effect graphs are illustrated on Figure~\ref{fig:argument-effects}. The
only node with write effects is \texttt{bar} that modifies \texttt{n} through a
subscript. Outgoing edges (1) and (2) can be removed because they are
outgoing edges of \texttt{bar} starting by \texttt{n}, which is already locally written by bar. Incoming edge (0) ends
by \texttt{n} that is written by \texttt{bar}, so it propagates the write
effects to the source node corresponding to the function \texttt{root}. After
removing this edge, the only remaining edges are propagating reads and the
analysis ends. The analysis informs that \texttt{root} updates \texttt{r}
and \texttt{bar} updates \texttt{n}, but \texttt{foo} doesn't update its
argument.

@@

            \draw[->] (bar) -- node[right] {$(\texttt{n,s})$} node[left] {(2)} (foo);
            \draw[->] (bar.east) node[right, yshift=-1em] {(1)} -| ($(bar.east) + (.5,.75)$) node[right, yshift=-.8em] {$(\texttt{n,n})$} -| ($(bar.north) + (.5,0)$) ;
            \draw[->] (root) -- node[right, yshift=.8em] {$(\texttt{r,n})$} node[left, yshift=.8em] {(0)} (bar);
        \end{tikzpicture}
        \caption{Initial argument effect graph..}
        \label{fig:argument-effects-graph}
    \end{subfigure}

@@

A naive translation of such a call would generate a
call to a C++ function that would allocate enough memory to hold three integers,
fill the corresponding buffer and return it. A smarter version would generate a
static array holding the relevant values. Figure~\ref{fig:unroll-range}
compares two translations of a Python snippet that iterates over a fixed-size
list. The Clang compiler successfully turns Listing~\ref{lst:stat} into a
single constant. It also statically computes the return value for
Listing~\ref{lst:dyna}, but still generates the initialization code for the
\texttt{std::vector}. The GCC compiler fails to statically compute the return
value in both cases. This example shows that constant propagation should be
done at a higher level than C++, and so be part of the Python optimizer.

@@

Once interprocedural constant evaluation is done (eventually followed by
forward substitution), iteration over fixed size lists may be exposed, giving an
opportunity to fully unroll these loops at the Python level, thus avoiding the
suboptimal C++ translation observed for Listing~\ref{lst:stat}
and~\ref{lst:dyna} only.

@@

By construction of the expression template, the \texttt{eval} method used in
\texttt{from.eval(i)} does not have side effects, and the Python to C++
compiler can assert there is no shared memory between \texttt{to} and any
component of \texttt{from}. In that case, all loops can be made parallel, and
the innermost loop may be vectorized.

@@

A conservative approach is to start a parallel region if the loop trip
count is large enough. We avoid using the \texttt{if} clause for parallel
regions in order to prevent the unnecessary overhead. Instead a manual
\texttt{if} is used: the loop statement is duplicated and guarded by the appropriate
check, and only one loop holds the parallel annotations.
%\todo{this paragraph seems to be about a particular runtime issue, this shouldn't matter.
%Moreover, how is it relevant to this paper? It is an "implementation detail"}

Note that this parallelization scheme is compatible with the manual
parallelization of Python programs using OpenMP directives ported to the Python
language as proposed in~\cite{pyhpc2013}.
%\todo{I think a specific section could be made on this topic} SG: I lack time

@@

@@

As introduced in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test
suite to evaluate the optimizations described in\S~\ref{sec:pythran}.
Pythran (commit: 488b7ba) is used  with and without SIMD optimization.
The results are compared with Pypy (commit: 632c832bd32f), Parakeet (commit: 4e55810),
Numba (commit: b8a79c5) and Python 2.7.

The evaluation process consists in  11 runs for each benchmark. The execution
time of a single run is around 1 second.  The best time, average time and
standard deviation are computed from these sequences.

Benchmarking is done on two architectures with different back-end compilers.

The first one holds 8 AMD Magny Cours processors for a
total of 48 cores. Each core has access to 64\,KB of L1 cache,
512\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 6 cores of a processor. This configuration
provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
up to SSE4. We refer to this configuration as \textbf{AMD48}.
On AMD48, the back-end compiler is the clang-omp
version from Intel\footnote{\url{http://clang-omp.github.io/}} to be able to
benefit from OpenMP improvements. This Clang version is based on Clang 3.4
release (LLVM:233b1e3f034, clang-omp:6f358bcd87).

The second configuration holds one Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
8 cores (multithreading is not used). Each node has access to 64\,KB of L1
cache, 256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 8 cores. This configuration provides a total of 64\,GB of 
main memory. It supports up to AVX. We refer to this configuration as
\textbf{Intel8}.
On Intel8, backend compilers is GCC 4.9 (20140528) with libgomp.

@@

% For performance comparison between different Pythran version
To compare the different optimization schemes proposed in \S~\ref{sec:pythran} and their effect with respect to the leading example from \S~\ref{sec:optimize}, we evaluated different optimization combination for Listing~\ref{lst:rosen}. The input is a raw array of 1,000,000 single precision floating point elements.
Figures~\ref{fig:rosenamd48} and \ref{fig:rosenintel8} illustrate the results.
The speedup difference between the two architectures are explained by AVX support on
the Intel8 architecture while AMD48 supports up to SSE4 only.
We can also notice the importance of forward substitution on this benchmark as
we avoid one extra loop, two temporary array assignments and some load/store for
SIMD instructions.

@@

\subsection{Benefits of Implicit Parallelism}

The \texttt{arc\_distance} benchmark is a typical example for which we can make
intensive use of the techniques described in this paper. This kernel performs
intensive points to points computation and trigonometric use.
Figures~\ref{fig:arc_distance} and~\ref{fig:arc_distance_intel8} show speedups
compared to one thread execution. We can see that OpenMP improvement is
scalable up to 20 threads with a raw array of 10,000,000 double precision
floating point elements as input.

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{arc_distance_amd48.png}
\caption{Speed up for the \texttt{arc\_distance} kernel on AMD48.}
\label{fig:arc_distance}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{arc_distance_intel8.png}
\caption{Speed up for the \texttt{arc\_distance} kernel on Intel48.}
\label{fig:arc_distance_intel8}
\end{figure}

%Finally, Figure~\ref{fig:l2norme} and \ref{fig:l2norm_intel8} illustrate the
%execution time for the \texttt{l2norm} kernel introduced in
%\S~\ref{sec:stackoverflow}. Comparison with other compilers shows that the
%approach taken by Pythran is relevant for these kind of functions. Just like
%for the \texttt{arc\_distance} benchmark, we notice the benefits of using the
%AVX instruction set when available.
%
%\begin{figure}[t]
%\centering
%\includegraphics[width=0.5\textwidth]{l2norm_amd48.png}
%\caption{Execution time of the l2norm kernel on AMD48.}
%\label{fig:l2norme}
%\end{figure}
%
%\begin{figure}[t]
%\centering
%\includegraphics[width=0.5\textwidth]{l2norm_intel8.png}
%\caption{Execution time of the l2norm kernel on Intel8.}
%\label{fig:l2norm_intel8}
%\end{figure}

@@

\begin{table*}[t]

@@

The authors are very grateful to Béatrice Creusillet for her careful proof reading.
